### 2020-09-07 manuscript.tex
typo

With the overwhelming availability of biological sequence data, the development of efficient Bayesian phylogenetic methods is more important than ever. 
The performance of MCMC is dependent not only on computational performance but also the efficacy of an
MCMC setup to achieve its convergence.
A critical task therein lies the further advancement of MCMC operators.
Recent developments in this area include the advancement of guided tree proposals \cite{zhang2020using, meyer2019adaptive, hohna2012guided}, coupled MCMC \cite{altekar2004parallel, muller2019coupled},
adaptive multivariate transition kernels \cite{baele2017adaptive},
and other explorative proposal kernels such as the Bactrian and mirror kernels \cite{yang2013searching, thawornwattana2018designing}.
In the case of relaxed clocks, informed tree proposals can account for correlations between substitution rates and divergence times \cite{zhang2020improving}. 
The rate parameterisation itself can also affect the ability to ``mix'' during MCMC \cite{drummond2006relaxed, li2012model, zhang2020improving}.

@@

Let $\mathcal{T}$ be a binary rooted time tree with $N$ taxa, and %. Let $L$ be the number of sites within the multiple sequence alignment $D$. 
$D$ data associated with the tips, like a multiple sequence alignment with $L$ sites, morphological data or geographic locations. 
The posterior density of a phylogenetic model is described by

\begin{eqnarray}
\label{eq:bayesian}
p(\mathcal{T}, \vec{\mathcal{R}}^{\,}, \sigma, \mu_C, \theta|D) \propto  p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \mu_C, \theta) \; p(\mathcal{T}|\theta) \;  p(\vec{\mathcal{R}}^{\,} | \sigma) \; p(\sigma) \; p(\mu_C) \; p(\theta).
\end{eqnarray}

\noindent
where $\sigma$ and $\mu_C$ represent clock model related parameters, and $p(\mathcal{T}|\theta)$ is the tree prior where $\theta$ describes further unspecified parameters. 
The tree likelihood $p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \mu_C,\theta)$ %is calculated by the tree-pruning algorithm \cite{felsenstein1981evolutionary}, where
has $\mu_C$ as the overall clock rate and
$\vec{\mathcal{R}}^{\,}$ is an abstracted vector of branch rates which is transformed into real rates by function $r(\vec{\mathcal{R}}^{\,})$. 
Branch rates have a mean of 1 under the prior to avoid non-identifiability with node heights and the clock rate $\mu_C$.
Three methods of representing rates as $\vec{\mathcal{R}}^{\,}$ are presented in \textbf{\nameref{sect:rateparams}}.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Branch rate parameterisations.} Top left: the prior density of a branch rate $r$ under a Log-Normal($-0.5\sigma^2, \sigma$) distribution (with its mean fixed at 1). 
The function for transforming $\mathcal{R}$ into branch rates $r(\mathcal{R})$ is depicted for \textit{real} (top right), \textit{cat} (bottom left), and \textit{quant} (bottom right).
For visualisation purposes, there are only 10 bins/pieces displayed, however in practice we use $2N-2$ bins for \textit{cat} and 100 pieces for \textit{quant}.
The first and final \textit{quant} pieces are equal to the underlying function however the pieces in between use linear approximations of this function. }
\label{fig:rateparams}
\end{figure}

@@

The family of constant distance operators (\texttt{ConstantDistance}, \texttt{SimpleDistance}, and \texttt{SmallPulley} \cite{zhang2020improving}) are best suited for larger datasets (or datasets with strong signal) where the likelihood distribution is peaked (\textbf{Fig \ref{fig:rateparams}}).
While simple one dimensional operators such as \texttt{RandomWalk} or \texttt{Scale} must make small steps in order to stay ``on the ridge'' of the likelihood function, the constant distance operators ``wander along the ridge''  by ensuring that genetic distances are constant after the proposal.

@@

 \texttt{Scale} and \texttt{CisScale} both operate on the clock model standard deviation $\sigma$ however they behave differently in the \textit{real} and \textit{quant} parameterisations (\textbf{Fig \ref{fig:clockSDoperators}}).
In \textit{real}, large proposals of $\sigma \rightarrow \sigma^\prime$ made by  \texttt{Scale} could incur large penalties in the clock model prior density $p(\vec{\mathcal{R}}^{\,}|\sigma^\prime)$ and thus may be rejected quite often. 
This led to the development of the fast clock scaler \cite{zhang2020improving} (herein referred to as \texttt{CisScale}).
This operator recomputes all branch rates $\vec{\mathcal{R}}^{\,} \rightarrow \vec{\mathcal{R}}^{\, \prime}$ such that their quantiles under the new clock model prior remain constant $p(\vec{\mathcal{R}}^{\,}|\sigma) = p(\vec{\mathcal{R}}^{\, \prime}|\sigma^\prime)$.
In contrast, a proposal made by \texttt{Scale} $\sigma \rightarrow \sigma^\prime$ under the \textit{quant} parameterisation implicitly alters all branch rates $r(\vec{\mathcal{R}}^{\,})$ while leaving the quantiles $\vec{\mathcal{R}}^{\,}$ themselves constant.
In contrast, by applying \texttt{CisScale} under \textit{quant}
all quantiles are recomputed $\vec{\mathcal{R}} \rightarrow \vec{\mathcal{R}}^\prime$  such that the respective branch rates are constant, i.e. $r(\vec{\mathcal{R}}^{\,}) = r(\vec{\mathcal{R}}^{\,\prime})$.
In summary, \texttt{Scale} and \texttt{CisScale} propose rates/quantiles in the opposite (trans) or same (cis) space that the clock model is parameterised under (\textbf{Fig \ref{fig:clockSDoperators}}).

@@

\noindent
where \texttt{RF} is the Robinson-Foulds tree distance \cite{robinson1981comparison}, and $|p|$ is the number of dimensions of numerical parameter $p$ (1 for $\sigma$, $2N-2$ for $\vec{\mathcal{R}}^{\,}$, and $2N-1$ for node heights $t$).
The remaining terms are trained during the second and third phases: the cumulative computational runtime spent on each operator $\mathbb{T}(\omega)$, the sample standard deviation $\sigma_p$ of each numerical parameter $p$, and the summed distances $\sum_x \mathbb{D}(x_p, x_p^\prime)$.

@@

Under \textbf{Equations \ref{eqn:adaptiveSampler} and \ref{eqn:distanceFunctions}}, operators which effect larger changes on the parameters of interest, in shorter runtime, are sampled with greater probabilities. 
Division of the squared distance by a parameters variance $\sigma^2_p$ enables comparison between different numerical parameters.

@@

The step size of a proposal kernel $q(x^\prime|x)$ should be such that the proposed state $x^\prime$ is sufficiently far from the current state $x$ to explore vast areas of parameter space, but not so large that the proposal is rejected too often \cite{roberts1997weak}. 
Operators which attain an acceptance probability of 0.234 are often considered to have arrived at a suitable midpoint between these two extremes \cite{bouckaert2019beast, roberts1997weak}.
The standard uniform distribution kernel has recently been challenged by the Bactrian kernels \cite{yang2013searching, thawornwattana2018designing}.
The (Gaussian) $\text{Bactrian}(m)$ distribution is defined as the sum of two normal distributions:

@@

\noindent
where $0 \leq m < 1$ describes the modality of the Bactrian distribution. When $m=0$, the Bactrian distribution is equivalent to a $\text{Normal}(0, 1)$ distribution. 
As $m$ approaches 1, the distribution becomes increasingly bimodal (\textbf{Fig \ref{fig:bactrian}}). 
Yang et al. 2013 \cite{yang2013searching} demonstrate that $\text{Bactrian}(m=0.95)$ yields a proposal kernel which traverses the posterior distribution more efficiently that the standard uniform kernel, by placing minimal probability on steps which are too small or too large.
In this case, a target acceptance probability of around 0.3 is optimal.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of \texttt{NarrowExchange} and \texttt{NarrowExchangeRate} operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axis corresponds to node height $t$. In the bottom figure, branch rates $r$ are indicated by line width and therefore genetic distances are equal to the width of each branch multiplied by its length. 
In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.
%RB: at this stage, D_AE and D_CE constraints have not been defined yet. Remove or reformulate or move to after definition?
}
\label{fig:narrowexchange}
\end{figure}

We hypothesised that if \texttt{NarrowExchange} was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal (parallel to the constant distance operators \cite{zhang2020improving}), then its ability to traverse the state space may improve.

@@

\begin{align}
	r_i > 0 \text{ and } {r_i}^\prime > 0 \text { for } i \in \{A,B,C,D\} \\
	\max\{t_B, t_C \} < {t_D}^\prime < t_E.
\end{align}

@@

The total space of NER operators is comprised of all possible subsets of distance constraints (i.e. $\{\},\{\mathcal{D}_{AB}\}, \{\mathcal{D}_{AC}\}, \dotso , \{\mathcal{D}_{AB}, \mathcal{D}_{AC}, \mathcal{D}_{AE}, \mathcal{D}_{BC}, \mathcal{D}_{BE}, \mathcal{D}_{CE} \}$) that are solvable. 
The simplest NER kernel -- the null operator denoted by NER$\{\}$ -- does not satisfy any distance constraints and is equivalent to \texttt{NarrowExchange}. 
To determine which NER variants have the best performance, we developed an automated pipeline for generating and testing these operators.

@@

\paragraph{2. Solving Jacobian determinants.} The determinant of the Jacobian matrix $J$ is required for computing the Green ratio of the proposal. $J$ is defined as

@@

Computing the determinant $|J|$ invokes standard analytical differentiation and linear algebra libraries of MATLAB. 
6 of the 54 solvable operators were found to have $|J|=0$, corresponding to irreversible proposals, and were discarded.

@@

\textbf{Fig \ref{fig:acceptanceRateScreening}} also revealed a cluster of NER variants which -- under the conditions of the simulation --  performed better than the null operator $\text{NER}\{\}$ around 25\% of the time and performed worse around 10\% of the time. 
One such operator was  $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$  (\textbf{Algorithm \ref{alg:NER2}}). 
This variant conserves the genetic distance between nodes $A$, $B$, $C$ and their grandparent $E$. 
Moreover, this operator performed well when branch rates have a large variance ($\sigma > 0.5$), corresponding to non clock-like data. 
On the other hand, the null operator $\text{NER}\{\}$ performed better on shorter sequences ($L < 1$kb) with weaker signal.  
Overall, $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ outperformed the standard \texttt{NarrowExchange} operator when the data was not clock-like and contained sufficient signal.

@@

\begin{align}
	x_i = \begin{cases} \log r_i \text{ for } \textit{real} \\
						\log \frac{q_i}{1 - q_i} \text{ for } \textit{quant}  \end{cases}
\end{align}

\noindent
where $r_i$ is a real rate and $q_i$ is a rate quantile. The AVMVN probability density is defined by

\begin{align}
	\mathcal{AVMVN}(x) =  \mathcal{MVN}\big(x, (1-\beta) \frac{\Sigma_N}{N} + \beta \frac{\mathbb{I}_N}{N} \big) ,
\end{align}

@@

In all phylogenetic analyses presented here, we use a Yule \cite{yule1925ii} tree prior $p(\mathcal{T}|\lambda)$ with birth rate $\lambda \sim \text{Log-Normal}(1,1.25)$. %RB: mean in real space? This is not the traditional parameterisation in the literature, and it was also mentioned earlier that the mean parameter is set such that the mean=1
The clock standard deviation has a $\sigma \sim \text{Gamma}(0.5396,0.3819)$ prior.
Datasets are partitioned into subsequences, where each partition is associated with a distinct HKY substitution model \cite{hasegawa1985dating}.
The transition-transversion ratio $\kappa \sim \text{Log-Normal}(1, 1.25)$, the four nucleotide frequencies $(f_A, f_C, f_G, f_T) \sim \text{Dirichlet}(10,10,10,10)$, and the relative clock rate $\mu_C \sim \text{Log-Normal}(-0.18, 0.6)$ are estimated independently for each partition.
%RB: again mean in real space for Log-Normal?
The operator scheme ensures that the clock rates $\mu_C$ have a mean of 1 across all partitions. 
This avoids non-identifiability with branch substitution rates.
To enable the rapid benchmarking of larger datasets we use BEAGLE for high-performance tree likelihood calculations \cite{ayres2012beagle} and coupled MCMC with four chains for efficient mixing \cite{muller2019coupled}.
The neighbour joining tree \cite{saitou1987neighbor} is used as the initial state in each MCMC chain.

\subsection*{Operator configurations} \label{sect:operatorSchemes}

@@

\subsection*{Assessment criteria and datasets}

% Cite HKY somewhere hasegawa1985dating

To avoid a cross-product explosion, the five targets for clock model improvement are evaluated sequentially in the following order: \textbf{\nameref{sect:adaptiveSampling}}, \textbf{\nameref{sect:rateparams}}, \textbf{\nameref{sect:randomwalks}}, \textbf{\nameref{sect:NER}}, and \textbf{\nameref{AVMVN_sect}}.
The four operators introduced in these sections are summarised in \textbf{Table \ref{table:newOperators}}.
The setting which is considered to be the best in each step is then incorporated into the following step. This protocol and its outcomes are summarised in \textbf{Fig \ref{fig:tournament}}.

@@

Methodologies are assessed according to the following criteria.

\textbf{1. Validation}. This is assessed by measuring the coverage of all estimated parameters in well-calibrated simulation studies. These are presented in \textbf{\nameref{sect:WCSS_appendix}} and give confidence operators are implemented correctly. \\

\textbf{2. Mixing of parameters}. Key parameters are evaluated for the number of effective samples generated per hour (ESS/hr).
These key parameters are the likelihood $\mathcal{L}$ and prior $p$ densities,  tree length $l$ (i.e. the sum of all branch lengths), mean branch rate $\bar{r}$, branch rate of all leaf nodes $r$, and relaxed clock standard deviation $\sigma$.
We also include the HKY substitution model term $\kappa$. 
The mixing of $\kappa$ should not be strongly affected by any of the clock model operators, and thus it serves as a positive control in each experiment. \\

Methodologies are benchmarked using one simulated and eight empirical datasets.
The latter were compiled \cite{lanfear2019Github} and partitioned \cite{lanfear2016partitionfinder} by Lanfear as ``benchmark alignments'' (\textbf{Table \ref{table:datasets}}).
Each methodology is benchmarked for million-states-per-hour using the Intel Xeon Gold 6138 CPU (2.00 GHz).
These terms are multiplied by the ESS-per-state across 20 replicates on the New Zealand eScience Infrastructure (NeSI) cluster to compute the total ESS/hr of each dataset under each setting.
All methodologies use identical models and operator configurations, except where a difference is specified.

@@

7  &  80  &  1 &  10.0  &  0.9 & 0.82 &  \textit{Simulated data}  \\

@@

\end{tabular}
\caption{Benchmark datasets, sorted in increasing order of taxa count $N$. Number of partitions $P$, total alignment length $L$, and number of unique site patterns $L_\text{unq}$ in the alignment are also specified.
Clock standard deviation estimates $\hat\sigma$ are of moderate magnitude, suggesting that most of these datasets are not clock-like.
}
\label{table:datasets}
\end{table}

\clearpage
\subsection*{Round 1: A simple operator-weight learning algorithm can improve performance}

@@

This experiment also revealed that the standard \texttt{Scale} operator was preferred over \texttt{CisScale} for the \textit{real} configuration. Averaged across all datasets, the learned weights behind these two operators were 0.47 and 0.03.
This is due to the computationally demanding nature of \texttt{CisScale} which invokes the i-CDF function.
In contrast, the performance of \texttt{Scale} and \texttt{CisScale} were more similar under the \texttt{quant} configuration and were weighted at 0.28 and 0.45.
For both \textit{real} and \textit{quant}, proposals which altered quantiles, while leaving the rates constant (\texttt{Scale} and \texttt{CisScale} respectively), were preferred.

Overall, the \texttt{AdaptiveOperatorSampler} operator will be included in all subsequent rounds in the tournament.

@@

This experiment showed that the \textit{real} parameterisation greatly outperformed \textit{cat} on most datasets and most parameters (\textbf{Fig \ref{fig:round2Results}}).
This disparity was strongest for long alignments.
In the most extreme case, leaf substitution rates $r$ and clock standard deviation $\sigma$ both mixed around $50\times$ faster on the 15.5 kb seed plant dataset (Ran et al. 2018 \cite{Ran_2018}) for \textit{real} than they did for \textit{cat}.
The advantages in using constant distance operators would likely be even stronger for larger $L$.
Furthermore, \textit{real} outperformed \textit{quant} on most datasets, but this was mostly due to the slow computational performance of \textit{quant} compared with real, as opposed to differences in mixing prowess (\textbf{Fig \ref{fig:round2Resultsb}}).
Irrespective of mixing ability, the adapt (\textit{real}) configuration had the best computation performance and generated samples 40\% faster than adapt (\textit{cat}) and 60\% faster than adapt (\textit{quant}).

Overall, we have determined that \textit{real}, and its associated operators, makes the best parameterisation covered here and will proceed to the next rounds of the benchmarking.

@@

\subsection*{Round 3: Bactrian proposal kernels are around 15\% more efficient than uniform kernels}

>>>>

With the overwhelming availability of biological sequence data, the development of efficient Bayesian phylogenetic methods is more important than ever. 
The performance of MCMC is dependent not only on computational performance but also the efficacy of an
MCMC setup to achieve convergence.
A critical task therein lies the further advancement of MCMC operators.
Recent developments in this area include the advancement of guided tree proposals \cite{zhang2020using, meyer2019adaptive, hohna2012guided}, coupled MCMC \cite{altekar2004parallel, muller2019coupled},
adaptive multivariate transition kernels \cite{baele2017adaptive},
and other explorative proposal kernels such as the Bactrian and mirror kernels \cite{yang2013searching, thawornwattana2018designing}.
In the case of relaxed clocks, informed tree proposals can account for correlations between substitution rates and divergence times \cite{zhang2020improving}. 
The rate parameterisation itself can also affect the ability to ``mix'' during MCMC \cite{drummond2006relaxed, li2012model, zhang2020improving}.

@@

Let $\mathcal{T}$ be a binary rooted time tree with $N$ taxa, and %. Let $L$ be the number of sites within the multiple sequence alignment $D$. 
data $D$ associated with the tips, such as a multiple sequence alignment with $L$ sites, morphological data or geographic locations. 
The posterior density of a phylogenetic model is described by

\begin{eqnarray}
\label{eq:bayesian}
p(\mathcal{T}, \vec{\mathcal{R}}^{\,}, \sigma, \mu_C, \theta|D) \propto  p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \mu_C, \theta) \; p(\mathcal{T}|\theta) \;  p(\vec{\mathcal{R}}^{\,} | \sigma) \; p(\sigma) \; p(\mu_C) \; p(\theta)
\end{eqnarray}

\noindent
where $\sigma$ and $\mu_C$ represent clock model related parameters, and $p(\mathcal{T}|\theta)$ is the tree prior where $\theta$ describes further unspecified parameters. 
The tree likelihood $p(D|\mathcal{T}, r(\vec{\mathcal{R}}^{\,}), \mu_C,\theta)$ %is calculated by the tree-pruning algorithm \cite{felsenstein1981evolutionary}, where
has $\mu_C$ as the overall clock rate and
$\vec{\mathcal{R}}^{\,}$ is an abstracted vector of branch rates which is transformed into real rates by function $r(\vec{\mathcal{R}}^{\,})$. 
Branch rates have a mean of 1 under the prior to avoid non-identifiability with the clock rate $\mu_C$.
Three methods of representing rates as $\vec{\mathcal{R}}^{\,}$ are presented in \textbf{\nameref{sect:rateparams}}.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/rateparameterisation.pdf}
\caption{\textbf{Branch rate parameterisations.} Top left: the prior density of a branch rate $r$ under a Log-normal($-0.5\sigma^2, \sigma$) distribution (with its mean fixed at 1). 
The function for transforming $\mathcal{R}$ into branch rates $r(\mathcal{R})$ is depicted for \textit{real} (top right), \textit{cat} (bottom left), and \textit{quant} (bottom right).
For visualisation purposes, there are only 10 bins/pieces displayed, however in practice we use $2N-2$ bins for \textit{cat} and 100 pieces for \textit{quant}.
The first and final \textit{quant} pieces are equal to the underlying function (solid lines) however the pieces in between use linear approximations of this function (dashed lines).}
\label{fig:rateparams}
\end{figure}

@@

The family of constant distance operators (\texttt{ConstantDistance}, \texttt{SimpleDistance}, and \texttt{SmallPulley} \cite{zhang2020improving}) are best suited for larger datasets (or datasets with strong signal) where the likelihood distribution is peaked (\textbf{Fig \ref{fig:rateparams}}).
While simple one dimensional operators such as \texttt{RandomWalk} or \texttt{Scale} must take small steps in order to stay ``on the ridge'' of the likelihood function, the constant distance operators ``wander along the ridge''  by ensuring that genetic distances are constant after the proposal.

@@

 \texttt{Scale} and \texttt{CisScale} both operate on the clock model standard deviation $\sigma$ however they behave differently in the \textit{real} and \textit{quant} parameterisations (\textbf{Fig \ref{fig:clockSDoperators}}).
In \textit{real}, large proposals of $\sigma \rightarrow \sigma^\prime$ made by  \texttt{Scale} could incur large penalties in the clock model prior density $p(\vec{\mathcal{R}}^{\,}|\sigma^\prime)$ and thus may be rejected quite often. 
This led to the development of the fast clock scaler \cite{zhang2020improving} (herein referred to as \texttt{CisScale}).
This operator recomputes all branch rates $\vec{\mathcal{R}}^{\,} \rightarrow \vec{\mathcal{R}}^{\, \prime}$ such that their quantiles under the new clock model prior remain constant $p(\vec{\mathcal{R}}^{\,}|\sigma) = p(\vec{\mathcal{R}}^{\, \prime}|\sigma^\prime)$.
In contrast, a proposal made by \texttt{Scale} $\sigma \rightarrow \sigma^\prime$ under the \textit{quant} parameterisation implicitly alters all branch rates $r(\vec{\mathcal{R}}^{\,})$ while leaving the quantiles $\vec{\mathcal{R}}^{\,}$ themselves constant.
Whereas, application of \texttt{CisScale} under \textit{quant} results in all quantiles being recomputed $\vec{\mathcal{R}} \rightarrow \vec{\mathcal{R}}^\prime$  such that their rates are constant, i.e. $r(\vec{\mathcal{R}}^{\,}) = r(\vec{\mathcal{R}}^{\,\prime})$.
In summary, \texttt{Scale} and \texttt{CisScale} propose rates/quantiles in the opposite (trans) or same (cis) space that the clock model is parameterised under (\textbf{Fig \ref{fig:clockSDoperators}}).

@@

\noindent
where \texttt{RF} is the Robinson-Foulds tree distance \cite{robinson1981comparison}, and $|p|$ is the number of dimensions of numerical parameter $p$ (1 for $\sigma$, $2N-2$ for $\vec{\mathcal{R}}^{\,}$, and $2N-1$ for node heights $t$).
The remaining terms are trained during the second and third phases: 
the sample standard deviation $\sigma_p$ of each numerical parameter of interest $p$,
the cumulative computational runtime spent on each operator $\mathbb{T}(\omega)$,  and the summed distances $\sum_x \mathbb{D}(x_p, x_p^\prime)$.

@@

Under \textbf{Equations \ref{eqn:adaptiveSampler} and \ref{eqn:distanceFunctions}}, operators which effect larger changes on the parameters of interest, in shorter runtime, are sampled with greater probabilities. 
Division of the squared distance by a parameter's sample variance $\sigma^2_p$ enables comparison between numerical parameters which exist in different spaces.

@@

The step size of a proposal kernel $q(x^\prime|x)$ should be such that the proposed state $x^\prime$ is sufficiently far from the current state $x$ to explore vast areas of parameter space, but not so large that the proposal is rejected too often \cite{roberts1997weak}. 
Operators which attain an acceptance probability of 0.234 are often considered to have arrived at a suitable midpoint between these two extremes \cite{bouckaert2019beast, roberts1997weak}.
The standard uniform distribution kernel has recently been challenged by the family of Bactrian kernels \cite{yang2013searching, thawornwattana2018designing}.
The (Gaussian) $\text{Bactrian}(m)$ distribution is defined as the sum of two normal distributions:

@@

\noindent
where $0 \leq m < 1$ describes modality. 
When $m=0$, the Bactrian distribution is equivalent to $\text{Normal}(0, 1)$. 
As $m$ approaches 1, the distribution becomes increasingly bimodal (\textbf{Fig \ref{fig:bactrian}}). 
Yang et al. 2013 \cite{yang2013searching} demonstrate that $\text{Bactrian}(m=0.95)$ yields a proposal kernel which traverses the posterior distribution more efficiently than the standard uniform kernel, by placing minimal probability on steps which are too small or too large.
In this case, a target acceptance probability of around 0.3 is optimal.

@@

\begin{figure}[!h]
\includegraphics[width=\textwidth]{Figures/NarrowExchange.pdf}
\caption{\textbf{Depiction of \texttt{NarrowExchange} and \texttt{NarrowExchangeRate} operators.} Proposals are denoted by $\mathcal{T} \rightarrow \mathcal{T}^\prime$. The vertical axes correspond to node heights $t$. In the bottom figure, branch rates $r$ are indicated by line width and therefore genetic distances are equal to the width of each branch multiplied by its length. 
In this example, the $\mathcal{D}_{AE}$ and $\mathcal{D}_{CE}$ constraints are satisfied.
%RB: at this stage, D_AE and D_CE constraints have not been defined yet. Remove or reformulate or move to after definition?
}
\label{fig:narrowexchange}
\end{figure}

We hypothesised that if \texttt{NarrowExchange} was adapted to the relaxed clock model by ensuring that genetic distances remain constant after the proposal (analogous to constant distance operators \cite{zhang2020improving}), then its ability to traverse the state space may improve.

@@

\begin{align}
	r_i, {r_i}^\prime > 0 \text { for } i \in \{A,B,C,D\} \\
	t_B, t_C  < {t_D}^\prime < t_E.
\end{align}

@@

The total space of NER operators consists of all possible subsets of distance constraints (i.e. $\{\},\{\mathcal{D}_{AB}\}, \{\mathcal{D}_{AC}\}, \dotso , \{\mathcal{D}_{AB}, \mathcal{D}_{AC}, \mathcal{D}_{AE}, \mathcal{D}_{BC}, \mathcal{D}_{BE}, \mathcal{D}_{CE} \}$) that are solvable. 
The simplest NER kernel -- the null operator denoted by NER$\{\}$ -- does not satisfy any distance constraints and is equivalent to \texttt{NarrowExchange}. 
To determine which NER variants have the best performance, we developed an automated pipeline for generating and testing these operators.

@@

\paragraph{2. Solving Jacobian determinants.} The determinant of the Jacobian matrix $J$ is required for computing the Green ratio of this proposal. $J$ is defined as

@@

Solving the determinant $|J|$ invokes standard analytical differentiation and linear algebra libraries of MATLAB. 
6 of the 54 solvable operators were found to have $|J|=0$, corresponding to irreversible proposals, and were discarded.

@@

\textbf{Fig \ref{fig:acceptanceRateScreening}} also revealed a cluster of NER variants which -- under the conditions of the simulation --  performed better than the null operator $\text{NER}\{\}$ around 25\% of the time and performed worse around 10\% of the time. 
One such operator was  $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$  (\textbf{Algorithm \ref{alg:NER2}}). 
This variant conserves the genetic distance between nodes $A$, $B$, $C$ and their grandparent $E$. 
This operator performed well when branch rates had a large variance ($\sigma > 0.5$), corresponding to non clock-like data. 
On the other hand, the null operator $\text{NER}\{\}$ performed better on shorter sequences ($L < 1$kb) with weaker signal.  
Overall, $\text{NER}\{\mathcal{D}_{AE}, \mathcal{D}_{BE}, \mathcal{D}_{CE}\}$ outperformed the standard \texttt{NarrowExchange} operator when the data was not clock-like and contained sufficient signal.

@@

\begin{align}
	x = \begin{cases} \log r \text{ for } \textit{real} \\
						\log \frac{q}{1 - q} \text{ for } \textit{quant}  \end{cases}
\end{align}

\noindent
where $r$ is a real rate and $q$ is a rate quantile. The AVMVN probability density is defined by

\begin{align}
	\mathcal{AVMVN}(\vec{x}) =  \mathcal{MVN}\big(\vec{x}, (1-\beta) \frac{\Sigma_N}{N} + \beta \frac{\mathbb{I}_N}{N} \big) ,
\end{align}

@@

In all phylogenetic analyses presented here, we use a Yule \cite{yule1925ii} tree prior $p(\mathcal{T}|\lambda)$ with birth rate $\lambda \sim \text{Log-normal}(1,1.25)$.
Here and throughout the article, a $\text{Log-normal}(a,b)$ distribution is parameterised such that $a$ and $b$ are the mean and standard deviation in log-space.
The clock standard deviation has a $\sigma \sim \text{Gamma}(0.5396,0.3819)$ prior.
Datasets are partitioned into subsequences, where each partition is associated with a distinct HKY substitution model \cite{hasegawa1985dating}.
The transition-transversion ratio $\kappa \sim \text{Log-normal}(1, 1.25)$, the four nucleotide frequencies $(f_A, f_C, f_G, f_T) \sim \text{Dirichlet}(10,10,10,10)$, and the relative clock rate $\mu_C \sim \text{Log-normal}(-0.18, 0.6)$ are estimated independently for each partition.
The operator scheme ensures that the clock rates $\mu_C$ have a mean of 1 across all partitions. 
This avoids non-identifiability with branch substitution rates.
To enable rapid benchmarking of larger datasets we use BEAGLE for high-performance tree likelihood calculations \cite{ayres2012beagle} and coupled MCMC with four chains for efficient mixing \cite{muller2019coupled}.
The neighbour joining tree \cite{saitou1987neighbor} is used as the initial state in each MCMC chain.

@@

To avoid a cross-product explosion, the five targets for clock model improvement were evaluated sequentially in the following order: \textbf{\nameref{sect:adaptiveSampling}}, \textbf{\nameref{sect:rateparams}}, \textbf{\nameref{sect:randomwalks}}, \textbf{\nameref{sect:NER}}, and \textbf{\nameref{AVMVN_sect}}.
The four operators introduced in these sections are summarised in \textbf{Table \ref{table:newOperators}}.
The setting which was considered to be the best in each step was then incorporated into the following step. 
This protocol and its outcomes are summarised in \textbf{Fig \ref{fig:tournament}}.

@@

Methodologies were assessed according to the following criteria.

\textbf{1. Validation}. This was assessed by measuring the coverage of all estimated parameters in well-calibrated simulation studies. 
These are presented in \textbf{\nameref{sect:WCSS_appendix}} and give confidence operators are implemented correctly. \\

\textbf{2. Mixing of parameters}. Key parameters were evaluated for the number of effective samples generated per hour (ESS/hr).
These key parameters were the likelihood $\mathcal{L}$ and prior $p$ densities,  tree length $l$ (i.e. the sum of all branch lengths), mean branch rate $\bar{r}$, branch rate of all leaf nodes $r$, and relaxed clock standard deviation $\sigma$.
We also included the HKY substitution model term $\kappa$. 
The mixing of $\kappa$ should not be strongly affected by any of the clock model operators, and thus it served as a positive control in each experiment. \\

Methodologies were benchmarked using one simulated and eight empirical datasets.
The latter were compiled \cite{lanfear2019Github} and partitioned \cite{lanfear2016partitionfinder} by Lanfear as ``benchmark alignments'' (\textbf{Table \ref{table:datasets}}).
Each methodology was benchmarked for million-states-per-hour using the Intel Xeon Gold 6138 CPU (2.00 GHz).
These terms were multiplied by the ESS-per-state across 20 replicates on the New Zealand eScience Infrastructure (NeSI) cluster to compute the total ESS/hr of each dataset under each setting.
All methodologies used identical models and operator configurations, except where a difference is specified.

@@

7  &  80  &  1 &  10.0  &  0.9 & 0.82 &  Simulated data  \\

@@

\end{tabular}
\caption{Benchmark datasets, sorted in increasing order of taxon count $N$. Number of partitions $P$, total alignment length $L$, and number of unique site patterns $L_\text{unq}$ in the alignment are also specified.
Clock standard deviation estimates $\hat\sigma$ are of moderate magnitude, suggesting that most of these datasets are not clock-like.
}
\label{table:datasets}
\end{table}

\clearpage
\subsection*{Round 1: A simple operator-weight learning algorithm greatly improved performance}

@@

This experiment also revealed that the standard \texttt{Scale} operator was preferred over \texttt{CisScale} for the \textit{real} configuration. Averaged across all datasets, the learned weights behind these two operators were 0.47 and 0.03.
This was due to the computationally demanding nature of \texttt{CisScale} which invokes the i-CDF function.
In contrast, the performance of \texttt{Scale} and \texttt{CisScale} were more similar under the \textit{quant} configuration and were weighted at 0.28 and 0.45.
For both \textit{real} and \textit{quant}, proposals which altered quantiles, while leaving the rates constant (\texttt{Scale} and \texttt{CisScale} respectively), were preferred.

Overall, the \texttt{AdaptiveOperatorSampler} operator was included in all subsequent rounds in the tournament.

@@

This experiment showed that the \textit{real} parameterisation greatly outperformed \textit{cat} on most datasets and most parameters (\textbf{Fig \ref{fig:round2Results}}).
This disparity was strongest for long alignments.
In the most extreme case, leaf substitution rates $r$ and clock standard deviation $\sigma$ both mixed around $50\times$ faster on the 15.5 kb seed plant dataset (Ran et al. 2018 \cite{Ran_2018}) for \textit{real} than they did for \textit{cat}.
The advantages in using constant distance operators would likely be even stronger for larger $L$.
Furthermore, \textit{real} outperformed \textit{quant} on most datasets, but this was mostly due to the slow computational performance of \textit{quant} compared with real, as opposed to differences in mixing prowess (\textbf{Fig \ref{fig:round2Resultsb}}).
Irrespective of mixing ability, the adapt (\textit{real}) configuration had the best computational performance and generated samples 40\% faster than adapt (\textit{cat}) and 60\% faster than adapt (\textit{quant}).

Overall, we determined that \textit{real}, and its associated operators, made the best parameterisation covered here and it proceeded to the following rounds of benchmarking.

@@

\subsection*{Round 3: Bactrian proposal kernels were around 15\% more efficient than uniform kernels}

We benchmarked the adapt (\textit{real}) configuration with a) standard uniform proposal kernels, and b) Bactrian(0.95) kernels \cite{yang2013searching}.
These kernels applied to all clock model operators (\textbf{Table \ref{table:bactriankernels}}).
These results confirmed that the Bactrian kernel yields faster mixing than the standard uniform kernel (\textbf{Fig \ref{fig:round3Results}}).
All relevant continuous parameters considered had an ESS/hr, averaged across the 9 datasets, between 15\% and 20\% faster compared with the standard uniform kernel.
Although the Bactrian proposal made little-to-no difference to the caterpillar dataset (Kawahara et al. 2013 \cite{Kawahara_2013}), every other dataset did in fact benefit.
Bactrian proposal kernels proceeded to round 4 of the relaxed clock model optimisation protocol.
